Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-04-10 paper/paper.tex
Minor edits.

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
What is a developer's contribution to a repository? By only counting commits and number of lines changed, existing tools that analyze source code repositories fall short on showing the effective contributions made by each developer. When many commits are viewed as a group, the details are lost. Commit information can be misleading since lines of code give no indication of what was actually being worked on without careful examination of the changed code. Providing a semantic view of this information could provide deeper insights into how software projects evolve since changes to design and features are not clearly visible from line changes alone. We present TypeV: a method for visualizing Java source code repositories. Instead of counting line changes in a commit we extract detailed type information over time by using the differences between abstract syntax trees (ASTs). We are then able to track the additions and deletions of declarations and invocations for each type. Furthermore, we can track each author’s type usage over time. Using this tool we examine specific cases in well-known repositories where our tool reveals interesting and useful information. We then compare type coverage information from the AST compared to file coverage to determine if unique information is provided by type information. 
\end{abstract}

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=\textwidth]{antlr-3-code-freq-edit}
\caption{GitHub graph showing code frequency. Note that the graph was truncated to better show the details. Thus the changes that occur in mid 2015 are much larger than appear. The additions reach over 400k and the deletions go over 300k.}
\label{fig:GH_code_freq}
\end{figure*}

@@

The history and evolution of a software project is often tracked by version control systems, such as Git, in a programming language agnostic way. Updates to source code, documentation, and tests are recorded in the form of commits, which show the differences in lines between each subsequent revision of the software. Version control systems become vital for large projects since individual changes would be impossible to track manually. Large projects have thousands of individual changes made by many different developers. With more changes to peruse, it becomes more difficult to see how a project is evolving over time since there is so much information to look through. Most hosting services for Git repositories provide some basic tools for analysing projects over time~\cite{github-graphs,bitbucket-graphs}; but, as we will show, these tools often lack the kind of detail to gather insightful analysis. While commits may store individual changes to lines of code, they represent the change in \emph{structure} of the software.  By visualizing lines changed, or files changed, the current state-of-the-art for software project visualization rarely capture the \emph{meaning} behind each commit; lines of code alone does not describe how a repository changed at a particular point in time. Thus, information regarding contributions from developers is often misleading since the number of commits and lines of code changed does not necessarily reflect meaningful changes to a repository~\cite{robles2014}.

@@

In this paper, we propose a method that analyses the changes in type usage and method invocations over time, as well as tool to visualize this analysis, which we call TypeV. We claim that type usage offers more insight than line-based, file-based, and commit-based methods of visualization and analysis. Of the services we have surveyed, none of the tools visualize project evolution using granular type analysis per author via commit-by-commit abstract syntax tree differencing. For simplicity, we decided to only analyze Java projects, as it allowed us to quickly develop and test our software. Since our approach described in this paper can be generally applied to any statically-typed language, we plan to extend it to other languages.

In Java, a type can be a primitive, provided by the language itself, or object references. Primitive types are those that are built into the language such as an \texttt{int} or \texttt{char}. Object references point to instances of classes. Classes are defined in the project itself or through an included library. The types in a project represent the functionality available for a developer to work with. Java’s syntax encourages the creation of classes, thus we expect substantial Java projects to define many domain-specific types. Different parts of a program use different types and external dependencies. When writing or contributing to a project, the author’s choice of which part of the program to edit will in turn affect what types they end up working with --- that is, it will affect the author’s \emph{type coverage}.

Type information is useful for seeing what libraries and functions are being used.  For example when a part of code becomes \emph{deprecated}, continuing its use may lead to design flaws or even security vulnerabilities. For example, in Apache Nifi, an earlier version of the software used a compromised key-derivation functions; hence, this class was deprecated, but left in for compatibility with older versions of the software~\cite{nifi}. Viewing the type information of a change can show when usage of deprecated classes are added or removed from a project as well as how many remain over time. In a more general sense, we can see when types are being added or removed from the project as well as what types are heavily used throughout the project. Even better, we can see \emph{who} uses which types heavily.

Having tools for analysing repositories based on type usage and method invocation would benefit many different parts of the software development process. Software management requires a high level understanding of the state of a repository and the changes being made which affect the functionality of the software. Tracking developer contributions in a less misleading way is also important for management since important contributions may be small and focused, but nevertheless, significant. Line difference (or simply, \emph{diff}) only provide meaningful information when they are analysed by someone with familiarity with the source code being changed. Providing tools which bring more meaning to line differences would allow people less familiar with the code to understand changes being made. When large changes are made that cover many aspects of a project, it would also be useful to see how the change affects the software rather than a several thousand line diff.  Seeing how a project has changed over time is also vital for management since evaluating the state of completion of software can be very difficult.

In the following sections of the paper we will introduce a tool, TypeV, that we developed that visualizes changes to types used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differencing, which we describe in section~\ref{sec:methodology}. We seek to answer the following questions: \\ \\
\textbf{RQ1:} How can we extract type and method usage over time from repositories \\ \\
\textbf{RQ2:} Does the type information give any interesting or useful insight into the project? \\ \\
\textbf{RQ3:} How does type coverage compare to file coverage?

@@

In another paper, Wu \textit{et al.} \cite{wu2004} the Evolutionary Spectrogram is used but with greater control of granularity for files and software subsystems such as directory structure. Spectrograms are also used to display the cardinality of a developer’s commits. The cardinality of a commit is the number of files or subsystems it affects in a system. All of the developers changes together show how many changes affect broad sections of the software.

@@

The Java ASTs in our analysis were generated using Spoon~\cite{pawlak:hal-01169705}. Spoon is a tool for transforming and analysing Java source code. It breaks code up into a meta-model including the AST itself. The model consists of three parts: structural elements, code elements and references to program elements. ''The structural part contains the declarations of the program elements, such as interface, class, variable, method, annotation, and enum declarations.  The code part contains the executable Java code, such as the one found in method bodies.  The reference part models the references to program elements (for instance a reference to a type)'' \cite{pawlak:hal-01169705} The Spoon model is convenient because it provides the infrastructure for performing AST differencing (described later) while retaining the code elements (such as type information) for analysis afterwards. Spoon also preserves all type and package (library) information—information vital to our semantic analysis.

@@

\begin{figure}[!h]
\centering
\includegraphics[height=2.5in]{commits_view}
\caption{Commits increase in time from bottom up. Therefore, ASTs from B and C are diff’d against ASTs in A. ASTs from E are diff’d against D. Since D is a merge commit, its changes are ignored as they are recorded in C and B.}
\label{fig:commits}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\columnwidth]{context}
\caption{}
\label{fig:context}
\end{figure}

\section{The Tool}

@@

The analyzed data would be difficult to analyze if displayed as its raw text form; instead, we created an HTML5 web visualization application for intended for developers, project managers, and researchers to use (Figure~\ref{fig:main-view}).

TypeV provide a variety of views to help organize the data and provide a finer level of detail. The main view displays a series of stacked bar graphs which show changes to types over a period of time in a project (Figure~\ref{fig:main-view}). The blue and orange portions account for the number of additions and deletions in a similar fashion to GitHub’s code frequency graph (Figure~\ref{fig:GH_code_freq}). The height of each bar shows the number of changes to a type in a smaller unit of time. This is calculated by multiplying the bar height by the number of changes during that time period divided by the maximum number of changes attested for this type. A bar which fills the row vertically is the period with the largest number of changes to that type out of the visible time periods. The other bars in the row are proportional to that maximum number. The width of each bar represents a set time period and can be changed to an hour, day, week, or month. The rows are sorted so that the graph for the most used types appear the top and are in descending order.

When a user mouses over a specific bar in the graph they are given simple statistics for the type and time period at that location. This includes the fully-qualified name of the type, the number of commits, the number of authors, the number of additions, the number of deletions, and the start and end date of that time period. If the user clicks on the bar, they are given more detailed statistics which list the authors who made the commits and each commit message (Figure~\ref{fig:typev-deets}).

One issue with analyzing individual authors is that authors can commit from a variety of different computers which can have different emails and usernames attached to the git configuration. This means in the git log one author might appear as several different authors. There is no perfect way to deal with this problem; however, we allow the user of our software to select and view multiple users at a time. Thus allowing the user to view changes from an individual or group of developers.

@@

% use section* for acknowledgment
\section*{Acknowledgment}

The authors would like to thank...

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
What is a developer's contribution to a repository? By only counting commits and number of lines changed, existing tools that analyze source code repositories fall short on showing the effective contributions made by each developer. When many commits are viewed as a group, the details are lost. Commit information can be misleading since lines of code give no indication of what was actually being worked on without careful examination of the changed code. Providing a semantic view of this information could provide deeper insights into how software projects evolve since changes to design and features are not clearly visible from line changes alone. We present TypeV: a method for visualizing Java source code repositories. Instead of counting line changes in a commit we extract detailed type information over time by using the differences between abstract syntax trees (ASTs). We are then able to track the additions and deletions of declarations and invocations for each type. Furthermore, we can track each author's type usage over time. Using this tool we examine specific cases in well-known repositories where our tool reveals interesting and useful information. We then compare type coverage information from the AST compared to file coverage to determine if unique information is provided by type information. 
\end{abstract}

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=\textwidth]{antlr-3-code-freq-edit}
\caption{GitHub graph showing code frequency. Note that the graph was truncated to better show the details, thus the changes that occur in mid 2015 are much larger than shown. The additions reach over 400k and the deletions are over 300k.}
\label{fig:GH_code_freq}
\end{figure*}

@@

The history and evolution of a software project is often tracked by version control systems, such as Git, in a programming language agnostic way. Updates to source code, documentation, and tests are recorded in the form of commits, which show the differences in lines between each subsequent revision of the software. Version control systems become vital for large projects since individual changes would be impossible to track manually. Large projects have thousands of individual changes made by many different developers. With more changes to peruse, it becomes more difficult to see how a project is evolving over time since there is so much information to look through. Most hosting services for Git repositories provide some basic tools for analyzing projects over time~\cite{github-graphs,bitbucket-graphs}; however, as we will show, these tools often lack the kind of detail to gather insightful analysis. While commits may store individual changes to lines of code, they represent the change in \emph{structure} of the software.  By visualizing lines changed, or files changed, the current state-of-the-art for software project visualization rarely capture the \emph{meaning} behind each commit; lines of code alone does not describe how a repository changed at a particular point in time. Thus, information regarding contributions from developers is often misleading since the number of commits and lines of code changed does not necessarily reflect meaningful changes to a repository~\cite{robles2014}.

@@

In this paper, we propose a method that analyses the changes in type usage and method invocations over time, as well as a tool to visualize this analysis, which we call TypeV. We claim that type usage offers more insight than line-based, file-based, and commit-based methods of visualization and analysis. Of the services we have surveyed, none of the tools visualize project evolution using granular type analysis per author via commit-by-commit abstract syntax tree differencing. For simplicity, we analyze only Java projects, as it allowed us to quickly develop and test our software. Since our approach described in this paper can be generally applied to any statically-typed language, we plan to extend it to other languages.

In Java, a type can be a primitive, provided by the language itself, or object references. Primitive types are those that are built into the language such as an \texttt{int} or \texttt{char}. Object references point to instances of classes. Classes are defined in the project itself or through an included library. The types in a project represent the functionality available for a developer to work with. Java's syntax encourages the creation of classes, thus we expect substantial Java projects to define many domain-specific types. Different parts of a program use different types and external dependencies. When writing or contributing to a project, the author's choice of which part of the program to edit will in turn affect what types they end up working with --- that is, it will affect the author's \emph{type coverage}.

Type information is useful for seeing what libraries and functions are being used.  For example, when a part of code becomes \emph{deprecated}, continuing its use may lead to design flaws or even security vulnerabilities. In Apache Nifi, an earlier version of the software used a compromised key-derivation functions; hence, this class was deprecated, but left in for compatibility with older versions of the software~\cite{nifi}. Viewing the type information of a change can show when usage of deprecated classes are added or removed from a project as well as how many remain over time. In a more general sense, we can see when types are being added or removed from the project as well as what types are heavily used throughout the project. Even better, we can see \emph{who} uses which types heavily.

Having tools for analyzing repositories based on type usage and method invocation would benefit many different parts of the software development process. Software management requires a high level understanding of the state of a repository and the changes being made which affect the functionality of the software. Tracking developer contributions in a less misleading way is also important for management since important contributions may be small and focused, but nevertheless, significant. Line difference (or simply, \emph{diff}) only provide meaningful information when they are analyzed by someone with familiarity with the source code being changed. Providing tools which bring more meaning to line differences would allow people less familiar with the code to understand changes being made. When large changes are made that cover many aspects of a project, it would also be useful to see how the change affects the software rather than a several thousand line diff.  Seeing how a project has changed over time is also vital for management since evaluating the state of completion of software can be very difficult.

In the following sections of the paper we will introduce a tool, TypeV, that we developed that visualizes changes to types used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differencing, which we describe in section~\ref{sec:methodology}. We seek to answer the following questions: \\ \\
\textbf{RQ1:} How can we extract type and method usage over time from repositories? \\ \\
\textbf{RQ2:} Does the type information give any interesting or useful insight into the project? \\ \\
\textbf{RQ3:} How does type coverage compare to file coverage?

@@

In another paper, Wu \textit{et al.} \cite{wu2004} the Evolutionary Spectrogram is used but with greater control of granularity for files and software subsystems such as directory structure. Spectrograms are also used to display the cardinality of a developer's commits. The cardinality of a commit is the number of files or subsystems it affects in a system. All of the developers changes together show how many changes affect broad sections of the software.

@@

The Java ASTs in our analysis were generated using Spoon~\cite{pawlak:hal-01169705}. Spoon is a tool for transforming and analysing Java source code. It breaks code up into a meta-model including the AST itself. The model consists of three parts: structural elements, code elements and references to program elements. According to Spoon's authors: ''The structural part contains the declarations of the program elements, such as interface, class, variable, method, annotation, and enum declarations.  The code part contains the executable Java code, such as the one found in method bodies. The reference part models the references to program elements (for instance a reference to a type)''~\cite{pawlak:hal-01169705}. The Spoon model is convenient because it provides the infrastructure for performing AST differencing (described later) while retaining the code elements (such as type information) for analysis afterwards. Spoon also preserves all type and package (library) information—information vital to our semantic analysis.

@@

\begin{figure}[!h]
\centering
\includegraphics[height=2.5in]{commits_view}
\caption{Commits increase in time from bottom up. Therefore, ASTs from B and C are diff'd against ASTs in A. ASTs from E are diff'd against D. Since D is a merge commit, its changes are ignored as they are recorded in C and B.}
\label{fig:commits}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\columnwidth]{context}
\caption{Shows the pipeline for TypeV. First we list the commits in chronological order. Then we perform AST diffs for each file. Finally, we display the information.}
\label{fig:context}
\end{figure}

\section{TypeV: The Visualization Tool}

@@

The analyzed data would be difficult to analyze if displayed as its raw text form; instead, we created an HTML5 web visualization application which is intended for developers, project managers, and researchers to use (Figure~\ref{fig:main-view}). The source code is available online on GitHub,\footnote{\url{https://github.com/mdfeist/TypeV}}.

TypeV provide a variety of views to help organize the data and provide a finer level of detail. The main view displays a series of stacked bar graphs which show changes to types over a period of time in a project (Figure~\ref{fig:main-view}). The blue and orange portions account for the number of additions and deletions in a similar fashion to GitHub's code frequency graph (Figure~\ref{fig:GH_code_freq}). The height of each bar shows the number of changes to a type in a smaller unit of time. This is calculated by multiplying the bar height by the number of changes during that time period divided by the maximum number of changes attested for this type. A bar which fills the row vertically is the period with the largest number of changes to that type out of the visible time periods. The other bars in the row are proportional to that maximum number. The width of each bar represents a set time period and can be changed to an hour, day, week, or month. The rows are sorted so that the graph for the most used types appear the top and are in descending order.

When a user mouses over a specific bar in the graph they are given simple statistics for the type and time period at that location. This includes the fully-qualified name of the type, the number of commits, the number of authors, the number of additions, the number of deletions, and the start and end date of that time period. If the user clicks on the bar, they are given more detailed statistics which list the authors who made the commits and each commit message.

One issue with analyzing individual authors is that authors can commit from a variety of different computers which can have different emails and usernames attached to the Git configuration. This means in the \texttt{git log} one author might appear as several different users. There is no perfect way to deal with this problem; however, we allow the user of our software to select and view multiple users at a time. Thus allowing the user to view changes from an individual or group of developers.

@@

% use section* for acknowledgment
%\section*{Acknowledgment}
%The authors would like to thank...
